---
title: "Übung: 'Angewandte Datenverarbeitung und -analyse mit R'"
subtitle: "Methodenfortbildung der ASP"
description: "Exercises for Day 3&4 of the workshop"
author: "Maik Bieleke, PhD"
institute: "University of Konstanz"
date: 11/17/2024
date-format: long
# format: 
#   pdf:
#     toc: true
#     execute:
#       include: false
format: 
  html:
    toc: true
---

# Introduction

In this exercise, you will practice data processing and data analysis with R. The exercises are based on published data from this paper:

Wiewelhove, T., Raeder, C., Meyer, T., Kellmann, M., Pfeiffer, M., & Ferrauti, A. (2015). Markers for routine assessment of fatigue and recovery in male and female team sport athletes during high-intensity interval training. *PloS One*, *10*(10), e0139801. <https://doi.org/10.1371/journal.pone.0139801>

# Project Setup

@.  Create a new project in RStudio with folders for data and figures.
    
@.  Download the data from [this figshare repository](https://figshare.com/articles/dataset/_Markers_for_Routine_Assessment_of_Fatigue_and_Recovery_in_Male_and_Female_Team_Sport_Athletes_during_High_Intensity_Interval_Training_/1567810?file=2349341) and save the `.xlsx` file in the data folder of your project.

@.  Import the Excel file as data frame into R.

    ```{r}
    #| echo: true
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    # load the rio package
    library(rio)
    
    # relative path assumes that data folder is in project directory
    data <- import("data/pone.0139801.s001.xlsx")
    ```
    
# Data exploration

@.  Inspect the dataset: 

    - How many rows and columns does it have?
    - What are the column names?
    - What type of data is stored in each column?
    - Which values are stored?
    - Are there any missing values?
    - ...

    ```{r}
    #| echo: true
    #| eval: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    # one function that shows you all of this information is skimr::skim().
    skimr::skim(data)
    ```

# Data management

@.  Add a variable `ID` to the dataset that uniquely identifies each athlete.

    ```{r}
    #| echo: true
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    data$ID <- 1:nrow(data)
    ```

@.  Rename the column `Sex [1=male]` to `Sex` and convert it from a numeric variable to a factor variable (with 1 = male, 2 = female).

    ```{r}
    #| echo: true
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    # rename the column with dplyr::rename()
    data <- dplyr::rename(data, Sex = `Sex [1=male]`)
    
    # convert it to a factor with factor().
    data$Sex <- factor(data$Sex, levels = c(1, 2), 
                       labels = c("male", "female"))
    ```
    
@.  The `table()` function is useful to count the distinct values of a variable (we did not cover this function in the workshop). Use it to count the number of females and males in the dataset by applying it to the `Sex` column.

    ```{r}
    #| echo: true
    #| eval: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    table(data$Sex)
    ```  


    
# Data analysis

@.  Athletes rated the delayed onset muscle soreness (DOMS) on a visual analogue scale from 0 to 100 at all three measurement points (variables `DOMS_Pre`, `DOMS_Post1`, and `DOMS_Post2`). Compute means and standard deviations for all three measurement points (1) across all athletes and (2) separately for females and males.

    ```{r}
    #| echo: true
    #| eval: false
    #| warning: false
    #| message: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    library(dplyr)
    
    # across all athletes
    summarize(data, 
              across(c(DOMS_Pre, DOMS_Post1, DOMS_Post2), mean))
    summarize(data, 
              across(c(DOMS_Pre, DOMS_Post1, DOMS_Post2), sd))
    
    # separately by gender
    summarize(group_by(data, Sex), 
              across(c(DOMS_Pre, DOMS_Post1, DOMS_Post2), mean))
    summarize(group_by(data, Sex), 
              across(c(DOMS_Pre, DOMS_Post1, DOMS_Post2), sd))
    ```

@.  For an ANOVA analysis, we need the data in a long format. This can be achieved with the `pivot_longer()` function from the `tidyr` package. However, we need a [regular expression](https://en.wikipedia.org/wiki/Regular_expression) to separate the variable names into two columns (`names_sep = "_(?=[^_]+$)"`) because the variable names are irregular. We did not cover this in the workshop, but you can use the following code snippet to achieve this:

    ```
    library(tidyr)
    
    # use the pivot_longer() function as we have done in the workshop
    # the only difference is that we use the regular expression in names_sep
    # because the number of dashes varies in the variable names
    data_long <- tidyr::pivot_longer(data, cols = -c(ID, Sex), 
                                     names_sep = "_(?=[^_]+$)",
                                     names_to = c("Measurement", "Time"), 
                                     values_to = "Value")
    ```

    ```{r}
    #| echo: false
    #| warning: false
    #| message: false
    
    library(tidyr)
    data_long <- tidyr::pivot_longer(data, cols = -c(ID, Sex), 
                                     names_sep = "_(?=[^_]+$)",
                                     names_to = c("Measurement", "Time"), 
                                     values_to = "Value")
    ```

    Make sure that you understand how this works in principle. You might ask ChatGPT for a detailed explanation of the regular expression if you are interested.

@.  Based on `data_long`, conduct a mixed-effects ANOVA to examine the effect of `Sex` and `Time` on RPE measurements (`Session_RPE`) during the HIIT training .

    ```{r}
    #| echo: true
    #| warning: false
    #| message: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    library(afex)
    
    # the RPE measurements must be filtered from data_long for the analysis
    data_anova <- dplyr::filter(data_long, Measurement == "Session_RPE")
    
    # estimate the ANOVA model
    model <- afex::aov_4(Value ~ Sex + (Time | ID), data_anova)
    ```
    
    Note that the model automatically deletes participants with missing data.
    
@.  Follow up the significant main effect of `Time` with polynomial contrasts. First, use the `emmeans` package to estimate marginal means of `Time` from the ANOVA model and assign them to an object `emm`. Then, use `contrast(emm, "poly")` to get polynomial contrasts for the `Time` variable. Is there any evidence for a linear trend in the data?

    ```{r}
    #| echo: true
    #| warning: false
    #| message: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    library(emmeans)
    
    # estimate the marginal means based on the ANOVA model
    emm <- emmeans::emmeans(model, ~ Time)
    
    # use function provided in the instructions to estimate contrasts
    emm_contrasts <- contrast(emm, "poly")
    
    # linear contrast is the first row of the output and significant
    ```

# Data visualization

@.  Create a plot that visualizes the estimated marginal means of `Time` along with their 95% confidence intervals with *ggplot2*.

    - Convert the estimated means `emm` object to a data frame using the `as.data.frame()` function. Inspect the result to see in which columns the means and CIs are.
    - Use the `ggplot()` function to create a plot with `Time` on the x-axis and the estimated marginal means on the y-axis.
    - Add error bars to the plot using `geom_errorbar()`. Consult the documentation of `geom_errorbar()` to learn how to specify the `ymin` and `ymax` aesthetics. (Hint: Review the examples at the end of the documentation.)
    - Add points to the plot using `geom_point()`.
    - Change the theme to `theme_bw()`.
    - Label the x-axis as "Time", the y-axis as "Estimated marginal means of RPE", and the title as "Polynomial contrasts of RPE over time".
    - Save the plot as a PNG file in the figures folder of your project.

    ```{r}
    #| echo: true
    #| warning: false
    #| message: false
    #| code-line-numbers: false
    #| code-fold: true
    #| code-summary: Solution
    
    library(ggplot2)
    
    # convert estimated means object to a data frame
    emm.df <- as.data.frame(emm) # inspect the data frame
    
    # create the plot
    p <- ggplot(data = emm.df, aes(x = Time, y = emmean)) +
      theme_bw() +
      geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
      geom_point() +
      labs(x = "Time", y = "Estimated marginal means of RPE", 
           title = "Polynomial contrasts of RPE over time")
    
    # save the plot
    ggsave("figures/polynomial_contrasts.png", plot = p, 
           width = 6, height = 4, dpi = 300)
    ```
    
    Below you see the desired result.
    
    ```{r}
    #| echo: false
    #| include: true
    p
    ```
    
Congratulations, you made it to the end of the exercise! If you want, you can also try to reproduce the analyses from the paper.